Parallel Adaptive Importance Sampling

### 2015-08-05 PAIS.tex
Conclusions done
Conclusions done, now sending to Colin.

@@

  of processors. This demonstrates the added

@@

Many advances have been made in the field of MCMC, to design ever more
complex methods, which more intelligently propose moves which leads to
faster converging methods. Function space versions of standard methods
such as the random walk Metropolis-Hastings (RWMH) algorithm or the
Metropolis adjusted Langevin algorithm (MALA), whose convergence rates
are independent of dimension have been
developed\cite{cotter2013mcmc}. The hybrid (or Hamiltonian) Monte
Carlo (HMC) method uses Hamiltonian dynamics in order to propose and
accept moves to states which are a long way away from the current
position\cite{sexton1992hamiltonian}, and function space analogues of
this have also been proposed\cite{beskos2011hybrid}. Riemannian
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.  These
methods allow us to more fully explore the posterior distribution, at
the cost of fewer iterations of the method.

@@

In this paper, we present a framework for parallelisation of
importance sampling, which allows us to use any of the current
Metropolis-based methodologies in order to create an efficient target
proposal from the current state of all of the chains in the
ensemble. In Section \ref{sec:Prelim} we outline some preliminaries,
including the general set up for Bayesian inverse problems, the
preconditioned Crank-Nicolson Langevin (pCNL) algorithm and a
brief review of the optimal transport resampler,  both of which we will be
employing within the algorithm. In Section \ref{sec:MCMC}, we explore the common problems with MCMC
methods, which we intend to address with the development of our new
framework. We describe this framework in Section \ref{Sec:PAIS}. In
Section \ref{Sec:Num}, we present some numerical experiments which
demonstrate the savings that can be made by employing this approach as
opposed to a naive/trivial parallelisation of existing MCMC
methods. In Section \ref{Sec:Conc}, we will summarise our results and
suggest some areas for future investigation.

\section{Preliminaries}
\subsection{Bayesian inverse problems}
In this paper, we particularly focus on the use of MCMC methods in characterising
posterior probability distributions in Bayesian inverse problems. We
wish to learn about a particular unknown quantity $u$, of which we are
able to make direct or indirect noisy observations. For now
we say that $u$ is a member of a Hilbert
space $X$.

@@

\subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}
In recent years, work has been carried out to frame MCMC proposal
distributions on function space\cite{cotter2013mcmc}. These new
discretisations perform comparably with the original versions in low
dimensions, so we choose one of these distributions to consider. If gradient information regarding the observation operator is
available, then a range of MCMC methods are available which exploit this information to improve mixing rates. One
example of such an algorithm is MALA. In \cite{cotter2013mcmc}, a function space version of this
method was presented, the pCNL algorithm, and is described in full in Table
\ref{tab:pCNL}. The proposal used in this method comes about through
a Crank-Nicolson approximation of the Langevin SDE, whose invariant
measure is the posterior measure $\mu_Y$.

@@

Analysis of this method shows that as the ensemble size increases, the statistics of the evenly weighted sample approach those of the posterior distribution, at least well enough for a proposal distribution as described in Section~\ref{sec:PAIS_alg}. The histogram of the evenly weighted sample exhibits small oscillations in the tails of the posterior, and also struggles to deal with discontinuities.

@@

Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm loses its ability to accurately represent the mean of the posterior distribution as the dimension increases. Clearly for a simple Gaussian problem, we can not resample in much more than 10 dimensions and hope to recover a representative sample.

@@

However, we argue that this is a far from optimal
scenario. First of all, unless we have a lot of information about the posterior, we will begin the algorithm a long way from statistical equilibrium. Some initial iterations then are not samples from the posterior, and must be thrown away. This process is known as the burn-in. In a trivially parallelised scenario, each process must perform this process independently, while also tuning the scaling parameter.

@@

These deficiencies of the trivial method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(PAIS). In the next section we will introduce the method in its most
general form. We will then introduce the version that we have
implemented, which utilises the resampler recently suggested by
Reich\cite{reich2013nonparametric} described in Section \ref{sec:filters}, and the pCNL proposal distribution.

\section{The Parallel Adaptive Importance Sampler \allowbreak (PAIS)}\label{sec:PAIS_alg}

In a trivially parallelised MCMC scheme, if we have enough processes
all sampling from the target density at the same time, then we can
make a very rough approximation of that density by taking the current
states of all of the processes as a representative sample. In the
PAIS, the proposal density for the importance sampler stage is a
mixture of all of the proposal densities for a chosen MCMC scheme,
using the current states. If the proposal in the MCMC scheme is
appropriate and our number of samples is big enough, we might hope
that this mixture of densities is a good enough approximation of the
target density, and that new samples drawn from
the mixture proposal are approximately draws from the target
measure. If this is the case, then the variance of importance weights
for these samples will be small, leading to a highly efficient
sampler.

The problem at this stage, is that our current states all have varying
importance, but in order for us to iterate this whole process again,
we require them to be the same. Therefore, we employ a resampler to
take our new set of samples with differing weights, and return a set
of states with equal weights, which are chosen so that the correlation
between the two sets is maximised. At this stage we can start the whole
process again. The algorithm is given in more detail in Table \ref{tab:PAIS}.

We wish to sample states $x \in X$ from a posterior
probability distribution $\mu_Y$. Since we have $M$ processes, we
represent the current state of all of the Markov chains as a vector
$\X = [x_1,x_2,\ldots,x_M]^T$. We are also given a transition kernel
$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example.

@@

Since the resampling does not give us a statistically identical sample
to that which is inputted, we cannot assume that the samples $\X^{(i)}$
are samples from the posterior. Therefore, as with serial
importance samplers, the weighted samples $(\W^{(i)},\Y^{(i)})_{i=1}^N$ are stored.

The key here is to choose a suitable transition kernel $\nu$ such that
if $X^{(i)}$ is a decent respresentative sample of the posterior,
the mixture density $\chi(\cdot;\X^{(i)})$ is a reasonable
approximation of the posterior distribution. If this is the case,
the newly proposed states $\Y^{(i)}$ will also be a good (and relatively
independent) sample of the posterior with low variance in the
weights $\W^{(i)}$.

In Section \ref{sec:num}, we will demonstrate how the algorithm
performs, using pCNL algorithm. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
processes can improve on the original MCMC algorithm and lead to
convergence in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost.

\section{Automated tuning of Algorithm Parameters}\label{sec:adap_delta}

Efficient selection of scaling parameters in MCMC algorithms is critical to achieving optimal mixing rates and hence achieving fast convergence to the target density. One significant difficulty with this approach, as with many kernel
approximation approaches, is finding an appropriate $\nu$ such that
$\chi$ is a close approximation to the posterior density $\pi$ in the
right situation. If the proposal distribution is too overdispersed, then the algorithm will propose states a long way from the current state but there will be a relatively low acceptance rate, impacting the quality of inferences on the posterior. Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the space so the mixing rate, and hence convergence rate, will be slow. It is therefore necessary to find a proposal distribution which is slightly overdispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.

Proposal distributions which are slightly over-dispersed as described above, can be found by tuning the variance of the proposal distribution during the burn-in phase of the algorithm. Algorithms which use this method to find optimal proposal distributions are known as adaptive MCMC algorithms. Adaptive MCMC algorithms will converge to the stationary distribution $\pi(\cdot)$, irrespective of whether the adaptive parameter itself converges to an optimal value, under the following conditions\cite{roberts2007coupling,roberts2009examples}.

Condition 1: The adaptive parameter exhibits diminishing adaptation, which says that the amount of movement in the adaptive parameter decreases with the length of the chain.

Condition 2: The smallest number of steps for which the kernel has sufficiently converged, from an initial state $x$, is finite, by bounded covergence.

Adaptively choosing $\delta$ with a large initial guess allows us to first explore the state space, searching for multiple modes, then reducing $\delta$ to an optimal value for efficient sampling helps combat these problems. This process is effective in the PAIS algorithm since tuning $\delta$ using all of the chains is faster than in a single chain. The resampling step causes the processes to spread out across the regions of high density so that when $\delta$ is reduced for locally efficient sampling, it is not required for the chains to constantly search out new modes.

@@

These statistics are usually not practical for finding optimal values of $\delta$ since they require knowledge of the analytic solution, and require that the algorithm be run for a long time to build up a sufficiantly large sample.

Another related statistic, is to use the variance of the estimate of the first moment $\hat{\mu}$. Assuming that the algorithm is converging to the invariant distribution, convergence will be fastest when the variance of the estimate is minimised. This fact follows from the convergence rate of Monte Carlo algorithms, $\sigma/\sqrt{n}$. The proposal distribution which minimises the variance, $\sigma^2$, of $\hat{\mu}$, is the most desirable. This variance statistic often converges faster than the moment itself, and does not require any knowledge of the true value of the moment.

@@

The effective sample size can be defined in the following way:
\[
	\neff = \frac{\left(\sum_{i=1}^M \! w_i\right)^2}{\sum_{i=1}^M \! w_i^2} \approx \frac{M\E(w)^2}{\E(w^2)} = M\left(1 - \frac{\mbox{var}(w)}{\mathbb{E}(w^2)}\right).
\]
The second two expressions are true when $M\rightarrow\infty$. From the last expression we see that when the variance of the weights is zero, which is our ideal scenario, $\neff = M$. Also, in a neighbourhood around $\delta_{\text{var}}^*$, $\neff$ decreases, when the variance increases. So if $v(w(y))$ can be equal to zero for a particular problem and proposal distribution, maximising the effective sample size is equivalent to minimising the variance of the weights. In the PAIS algorithm we have a dynamic proposal which is perturbed at every iteration. Hence, our proposal distribution cannot be exactly proportional to the posterior.

@@

If we are to use the effective sample size ratio, we need to look at how it behaves in different situations. In the next section we see how it behaves for different observation operators. Here we look at how the statistic behaves as we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results for the second problem discussed in \ref{sec:problem 2} when using the PAIS-pCNL algorithm to explore the posterior distribution. The left part of the figure shows that as the ensemble size increases, the value of $\delta$ which gives the optimal effective sample size ratio decreases. This is to be expected since we are trying to fit more kernels into the posterior distribution, so the variance of each kernel needs to be decreased.

@@

A popular method for an adaptive MCMC algorithm is to view the scaling parameter as a random variable which we can sample during the course of the MCMC iterations. However, it can be slow and we may need an uninformative prior. Alternatively, the parameter may be randomly sampled from some interval at various points during the chain, a benefit being that it allows some exploration of the state space but never converges to an optimal value of the scaling parameter. We choose to use a divide and conquer scheme which optimises the effective sample size. Some more sophisticated examples are described in \cite{roberts2009examples} and \cite{Ji2013adaptive}. The proposed algorithm, is not presented as an optimal strategy but as an example of the benefits of tuning the algorithm using the effective sample size statistic.

In the adaptive algorithm described in Table~\ref{tab:adapt} we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm. We must choose some sequence of iterations at which to update $\delta^{(j)}$, $\{n_k\}_{k=1}$, which can be decided using a sequence in which the terms grow exponentially further apart. The algorithm is given for the PAIS method but a similar method is used to adaptively calculate $\delta^*$ for the pCNL algorithm.

When implemented, there are a number of parameters which must be chosen to allow efficient tuning of $\delta$. Firstly, the initial value of $\delta$ should be chosen so that the chains spread out quickly across the state space. If we always choose $\delta^{(0)}=2$ the chains may miss the high density region completely, so a smaller value of $\delta$ must be chosen. This can be seen for example in the optimisation graph for the problem described in Section~\ref{sec:problem 1}, Figure~\ref{fig:MH1 opt delta}, where there is a sudden jump at $\delta\approx 0.1$.

Another choice to be made is which statistic $T(\delta)$ to optimise. In the examples which follow, we have optimised the naively parallelised pCNL algorithm using the optimal acceptance rate $\hat{\alpha} = 0.75$. This value was found by using the L2 error minimums to calculate $\delta^*$. This means that we are minimising the statistic
\[
	T_{\text{MH}}(\delta) = \left| \frac{N_{\text{acc}}(\delta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\delta)$ is the number of accepted moves and $N_{\text{total}}$ is the total number of samples produced. For the PAIS algorithm, we are maximising the effective sample size as discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\delta) = \neff(\delta)$.

In the numerics that follow in Section \ref{sec:num}, we utilise the
adaptive scheme in Table \ref{tab:adapt} for finding the optimum value of $\delta$ in
the proposal step in the PAIS.

\begin{table}
\begin{mdframed}
\begin{algorithmic}
\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
\STATE Choose $\delta^{(0)} \in (0, 2]$, $\delta_L^{(0)} = 0.99\delta^{(0)}$, $\delta_U^{(0)} = 1.01\delta^{(0)} \wedge 2$.
\FOR{$i=0,1,2, \ldots, N$}
\STATE Sample $y_j^{(i)}~\sim~\nu(\cdot;x_j^{(i)},\delta_L^{(i)})$ for $j=1, \ldots,M/2$.
\STATE Sample $y_j^{(i)}~\sim~\nu(\cdot;x_j^{(i)},\delta_U^{(i)})$ for $j=M/2+1, \ldots, M$.
\STATE $\chi(y;\X^{(i)}) = M^{-1}
\left(\sum_{j=1}^{M/2} \nu(y;x_j^{(i)},\delta_L^{(i)})+\sum_{j=M/2+1}^{M} \nu(y;x_j^{(i)},\delta_U^{(i)})\right)$.
\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
\IF{$i$ is in $\{n_k\}_{k=1}$}
	\STATE For $w_{kj} = w_j^{(k)}$, $N$ iterations since last update,
	\STATE $T_L = NM(\sum_{k=i-N}^N\sum_{j=1}^{M/2} w_{kj})^2/(\sum_{k=i-N}^N\sum_{j=1}^{M/2} w_{kj}^2)$.
	\STATE $T_U = NM(\sum_{k=i-N}^N\sum_{j=M/2+1}^M w_{kj})^2/(\sum_{k=i-N}^N\sum_{j=M/2+1}^M w_{kj}^2)$.
	\STATE $\delta^{(i+1)} = \delta^{(i)} - \Delta\D T$.
\ELSE
	\STATE $\delta^{(i+1)} = \delta^{(i)}$.
\ENDIF
\IF{$i < K$}
	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=1}^{M/2} \rightarrow (\frac{1}{M/2}, x_j^{(i+1)})_{j=1}^{M/2}$
	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=M/2+1}^M \rightarrow (\frac{1}{M/2}, x_j^{(i+1)})_{j=M/2+1}^M$
\ELSE
	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=1}^M \rightarrow (\frac{1}{M}, x_j^{(i+1)})_{j=1}^M$
\ENDIF
\ENDFOR
\end{algorithmic}
\end{mdframed}\caption{A pseudo-code representation of the adaptive
  PAIS algorithm.}
\label{tab:adapt}
\end{table}

\section{Numerical Examples}\label{sec:num}

@@

@@

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow a Gaussian distribution. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the resulting form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_\Sigma - \frac{1}{2}\big\|u\big\|^2_{\mathcal{T}}\right),
\end{equation}
where $\Sigma$ and $\mathcal{T}$ are the covariances of the observational noise and prior distributions respectively. In the numerics which follow, we choose $\Sigma = \sigma^2I_N$ and $\mathcal{T} = \tau^2I_d$ with $\tau^2 =2$ and $\sigma^2 = 0.1$, and we observe $u=-2.5$ with noise drawn from the prior, resulting in $D = -2.6738662$. These values result in a posterior density in which the vast majority of the density is contained inside the high density region of the prior. This means that it should be straightforward for the algorithm to find the stationary distribution. The Kullback-Leibler (KL) divergence for this problem is $D_{KL}(\mu_Y || \mu_0) = 2.67$.

@@

{\bf (1) Finding the optimal parameters}: To find the optimal parameters we choose 32 values of $\delta$ evenly spaced in $[10^{-5}, 2]$. We run the PAIS-pCNL algorithm for 10,000 iterations and pCNL for 100,000 iterations, each with $M=50$ processes. We took 32 repeats of both algorithms and then used the medians of the statistics to find the optimal parameters.

@@

\subsubsection{Numerical optimal values of $\delta$}\label{sec:Optimal values P1}

@@

>>>>

%test change

@@

  of processors, for low dimensional problems, and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of processors used.

@@

Many advances have been made in the field of MCMC, to design ever more
complex methods, which more intelligently propose moves which leads to
faster converging methods. Function space versions of standard methods
such as the random walk Metropolis-Hastings (RWMH) algorithm or the
Metropolis adjusted Langevin algorithm (MALA), whose convergence rates
are independent of dimension have been
developed\cite{cotter2013mcmc}. The hybrid (or Hamiltonian) Monte
Carlo (HMC) method uses Hamiltonian dynamics in order to propose and
accept moves to states which are a long way away from the current
position\cite{sexton1992hamiltonian}, and function space analogues of
this have also been proposed\cite{beskos2011hybrid}. Riemann
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.  These
methods allow us to more fully explore the posterior distribution, at
the cost of fewer iterations of the method.

@@

In this paper, we present a framework for parallelisation of
importance sampling, which allows us to use any of the current
Metropolis-based methodologies in order to create an efficient target
proposal from the current state of all of the chains in the
ensemble. In Section \ref{Sec:Prelim} we outline some preliminaries,
including the general set up for Bayesian inverse problems, the
preconditioned Crank-Nicolson Langevin (pCNL) algorithm and a
brief review of the optimal transport resampler,  both of which we will be
employing within the algorithm. We describe the Parallel Adaptive
Importance Sampler (PAIS) in Section \ref{Sec:PAIS}, and in Section
\ref{Sec:adapt} we describe how algorithmic parameters can be
automatically tuned to provide optimal convergence. In
Section \ref{Sec:Num}, we present some numerical experiments which
demonstrate the savings that can be made by employing this approach as
opposed to a naive/trivial parallelisation of existing MCMC
methods. In Section \ref{Sec:Conc}, we will summarise our results and
suggest some areas for future investigation.

\section{Preliminaries}\label{Sec:Prelim}
In this Section we will introduce preliminary topics and algorithms
that will be referred to throughout the paper.
\subsection{Bayesian inverse problems}
In this paper, we particularly focus on the use of MCMC methods in characterising
posterior probability distributions in Bayesian inverse problems. We
wish to learn about a particular unknown quantity $u$, of which we are
able to make direct or indirect noisy observations. For now
we say that $u$ is a member of a Hilbert
space $X$.

@@

\subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}\label{Sec:pCNL}
In recent years, work has been carried out to frame MCMC proposal
distributions on function space\cite{cotter2013mcmc}. These new
discretisations perform comparably with the original versions in low
dimensions. If gradient information regarding the observation operator is
available, then a range of MCMC methods are available which exploit this information to improve mixing rates. One
example of such an algorithm is MALA. In \cite{cotter2013mcmc}, a function space version of this
method was presented, the pCNL algorithm, and is described in full in Table
\ref{tab:pCNL}. The proposal used in this method comes about through
a Crank-Nicolson approximation of the Langevin SDE, whose invariant
measure is the posterior measure $\mu_Y$.

@@

Analysis of this method shows that as the ensemble size increases, the statistics of the evenly weighted sample approach those of the posterior distribution, at least well enough for a proposal distribution as described in Section~\ref{Sec:PAIS}. The histogram of the evenly weighted sample exhibits small oscillations in the tails of the posterior, and also struggles to deal with discontinuities.

@@

Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm
loses its ability to accurately preserve the mean of the posterior
distribution as the dimension increases. Clearly for a simple Gaussian
problem, we can not resample in much more than 10 dimensions and hope
to recover a representative sample, without hugely increasing the
number of particles in the ensemble. As the cost of the ETMC is
$\mathcal{O}(n^2)$, this does limit the dimension for which this
approach can be used.

@@

However, we argue that this is a far from optimal
scenario. First of all, unless we have a lot of information about the posterior, we will begin the algorithm a long way from statistical equilibrium. Some initial iterations then are not samples from the posterior, and must be thrown away. This process is known as the burn-in. In a trivially parallelised scenario, each process must perform this process independently.

@@

These deficiencies of the trivial method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(PAIS). In the next section we will introduce the method in its most
general form. We will then introduce the version that we have
implemented, which utilises the resampler recently suggested by
Reich\cite{reich2013nonparametric} described in Section
\ref{sec:filters}, and the pCNL proposal distribution described in
Section \ref{Sec:pCNL}

\section{The Parallel Adaptive Importance Sampler \allowbreak (PAIS)}\label{Sec:PAIS}

% In a trivially parallelised MCMC scheme, if we have enough processes
% all sampling from the target density at the same time, then we can
% make a very rough approximation of that density by taking the current
% states of all of the processes as a representative sample. In the
% PAIS, the proposal density for the importance sampler stage is a
% mixture of all of the proposal densities for a chosen MCMC scheme,
% using the current states. If the proposal in the MCMC scheme is
% appropriate and our number of samples is big enough, we might hope
% that this mixture of densities is a good enough approximation of the
% target density, and that new samples drawn from
% the mixture proposal are approximately draws from the target
% measure. If this is the case, then the variance of importance weights
% for these samples will be small, leading to a highly efficient
% sampler.

% The problem at this stage, is that our current states all have varying
% importance, but in order for us to iterate this whole process again,
% we require them to be the same. Therefore, we employ a resampler to
% take our new set of samples with differing weights, and return a set
% of states with equal weights, which are chosen so that the correlation
% between the two sets is maximised. At this stage we can start the whole
% process again.

Important samplers can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen from which we
can draw samples, and each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use the states of a
set of parallel MCMC chains to construct a proposal distribution which
will be as close as possible to the target density. Given enough
processors, the states of all of these chains at one point
in time may be reasonably representative sample of the target density.

The proposal distribution could be constructed in many different ways,
but we choose to use a mixture distribution, made up of MCMC proposals
(in this paper, specifically the pCNL proposal from Section \ref{Sec:pCNL}). Once the proposal is
constructed, we can sample a new set of states from the proposal
distribution, and each is assigned a weight given by the ratio of the target
density and the proposal mixture distribution density. Assuming that
our proposal distribution is a good one, then the variance of the
weights will be small, and we will have many useful samples. Finally, we
need to create a set of evenly weighted samples which best represent
this set of weighted samples. This is achieved by implementing the
ETMC algorithm. Once we once again have a set of evenly weighted
samples that we believe represents the target distribution well, we
can iterate the process once again. The algorithm is given in more detail in Table \ref{tab:PAIS}.

We wish to sample states $x \in X$ from a posterior
probability distribution $\mu_Y$. Since we have $M$ processes, we
represent the current state of all of the Markov chains as a vector
$\X = [x_1,x_2,\ldots,x_M]^T$. We are also given a transition kernel
$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
the pCNL proposal presented in Section \ref{Sec:pCNL}.

@@

Since the resampling does not give us a statistically identical sample
to that which is inputted, we cannot assume that the samples $\X^{(i)}$
are samples from the posterior. Therefore, as with serial
importance samplers, the weighted samples
$(\W^{(i)},\Y^{(i)})_{i=1}^N$, or their statistics, are stored.

The key here is to choose a suitable transition kernel $\nu$ such that
if $X^{(i)}$ is a decent representative sample of the posterior,
the mixture density $\chi(\cdot;\X^{(i)})$ is a reasonable
approximation of the posterior distribution. If this is the case,
the newly proposed states $\Y^{(i)}$ will also be a good (and relatively
independent) sample of the posterior with low variance in the
weights $\W^{(i)}$.

In Section \ref{Sec:Num}, we will demonstrate how the algorithm
performs, using pCNL algorithm. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
processes can improve on the original MCMC algorithm and lead to
convergence in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost.

\section{Automated tuning of Algorithm Parameters}\label{Sec:adapt}

Efficient selection of scaling parameters in MCMC algorithms is critical to achieving optimal mixing rates and hence achieving fast convergence to the target density. One significant difficulty is finding an appropriate $\nu$ such that
$\chi$ is a close approximation to the posterior density $\pi$. If the proposal distribution is too over-dispersed, then the algorithm will propose states a long way from the current state but there will be a relatively low acceptance rate, impacting the quality of inferences on the posterior. Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the space so the mixing rate, and hence convergence rate, will be slow. It is therefore necessary to find a proposal distribution which is slightly over-dispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.

Proposal distributions which are slightly over-dispersed as described
above, can be found by tuning the variance of the proposal
distribution $\nu$ during the burn-in phase of the
algorithm. Algorithms which use this method to find optimal proposal
distributions are known as adaptive MCMC algorithms. Adaptive MCMC
algorithms will converge to the stationary distribution $\pi(\cdot)$,
irrespective of whether the adaptive parameter itself converges to an
optimal value, under the following
conditions\cite{roberts2007coupling,roberts2009examples}: 
\begin{enumerate}
\item The adaptive parameter exhibits diminishing adaptation, which says that the amount of movement in the adaptive parameter decreases with the length of the chain. 
\item The smallest number of steps for which the kernel has sufficiently converged, from an initial state $x$, is finite, by bounded convergence.
\end{enumerate}

The variance of the proposal distributions that are used within the
proposal mixture distribution plays a key role in how well the
proposal distribution represents the target distribution, and
therefore how quickly the algorithm converges. A large over-dispersed
proposal distribution will sample a lot from the tails of the
distribution, but may also find other regions of high density with
respect to the target distribution. A very small variance in the
proposal distributions will lead to a very rough proposal mixture
distribution. In practise, we wish to find the happy medium in which
the proposal distribution is slightly more dispersed than the target
density, whilst also being a good representation of the regions which
have high density with respect to the target distribution.

Adaptively choosing the variance of the proposal distributions with a
large initial guess allows us to first explore the state space
globally, searching for multiple modes. Reducing the proposal variances to an
optimal value then allows us to explore each region efficiently. The
fact that we have multiple chains allows us to assess quickly and
effectively what the optimal variance of the proposal distributions
should be.

@@

These statistics are not practical for finding optimal values of
$\delta$ since they require knowledge of the analytic solution, and
require that the algorithm be run for a long time to build up a
sufficiently large sample. However they can be used to assess the
ability of other statistical measures to find the optimal proposal
variances in a controlled setting.

Another related statistic, is the variance of the sample estimate of the first moment $\hat{\mu}$. Assuming that the algorithm is converging to the invariant distribution, convergence will be fastest when the variance of the estimate is minimised. This fact follows from the convergence rate of Monte Carlo algorithms, $\sigma/\sqrt{n}$. The proposal distribution which minimises the variance, $\sigma^2$, of $\hat{\mu}$, is the most desirable. This variance statistic often converges faster than the moment itself, and does not require any knowledge of the true value of the moment.

@@

The effective sample size can be defined in the following way:
\[
	\neff = \frac{\left(\sum_{i=1}^M \! w_i\right)^2}{\sum_{i=1}^M \! w_i^2} \approx \frac{M\E(w)^2}{\E(w^2)} = M\left(1 - \frac{\mbox{var}(w)}{\mathbb{E}(w^2)}\right).
\]
The second two expressions are true when $M\rightarrow\infty$. From
the last expression we see that when the variance of the weights is
zero, $\neff = M$, which is our ideal scenario. In a neighbourhood around $\delta_{\text{var}}^*$, $\neff$ decreases, when the variance increases. So if $v(w(y))$ can be equal to zero for a particular problem and proposal distribution, maximising the effective sample size is equivalent to minimising the variance of the weights. In the PAIS algorithm we have a dynamic proposal which is perturbed at every iteration.

@@

If we are to use the effective sample size ratio as an indicator of
how to tune the proposal variance, we need to look at how it behaves
in different situations. In Section \ref{Sec:Num} we see how it behaves for different observation operators. Here we look at how the statistic behaves as we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results for the second problem discussed in \ref{sec:problem 2} when using the PAIS-pCNL algorithm to explore the posterior distribution. The left part of the figure shows that as the ensemble size increases, the value of $\delta$ which gives the optimal effective sample size ratio decreases. This is to be expected since we are trying to fit more kernels into the posterior distribution, so the variance of each kernel needs to be decreased.

@@

A popular approach for adaptive MCMC algorithms is to view the scaling
parameter as a random variable which we can sample during the course
of the MCMC iterations. However, it can be slow to converge to the
optimal value, and we may need an
uninformative prior for the scaling parameter. Alternatively, the parameter may be randomly sampled from some interval at various points during the chain, a benefit being that it allows some exploration of the state space but never converges to an optimal value of the scaling parameter. We choose to use a divide and conquer scheme which optimises the effective sample size. Some more sophisticated examples are described in \cite{roberts2009examples} and \cite{Ji2013adaptive}. The proposed algorithm, is not presented as an optimal strategy but as an example of the benefits of tuning the algorithm using the effective sample size statistic.

From hereon in, we will assume that we are using the proposal from the
pCNL algorithm given in Section \ref{Sec:pCNL}, which is given by:
\begin{equation}Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
2\delta\mathcal{C}\nabla \Phi(u)+
\sqrt{8\delta} W\right], \qquad W \sim \mu_0,
\end{equation}
where $X$ is our current state and $Y$ is our proposed state. The
scaling parameter $\delta$ dictates the variance in the proposal distribution.

 In the adaptive algorithm described in Table~\ref{tab:adapt} in
 Appendix \ref{Sec:App}, we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm. We must choose some sequence of iterations at which to update $\delta^{(j)}$, $\{n_k\}_{k=1}$, which can be decided using a sequence in which the terms grow exponentially further apart. The algorithm is given for the PAIS method but a similar method is used to adaptively calculate $\delta^*$ for the pCNL algorithm.

When implemented, there are a number of parameters which must be
chosen to allow efficient tuning of $\delta$. Firstly, the initial
value of $\delta$ should be chosen so that the chains spread out
quickly across the state space. However, if it is chosen to be too
large, then the method may become unstable and inefficient.

\section{Numerical Examples}\label{Sec:Num}

@@

Since we are comparing against trivially parallelised MCMC algorithms,
we also need to decide which statistic $T(\delta)$ to optimise for
these approaches. In the examples which follow, we have optimised the naively parallelised pCNL algorithm using the optimal acceptance rate $\hat{\alpha} = 0.75$. This value was found by using the L2 error minimums to calculate $\delta^*$. This means that we are minimising the statistic
\[
	T_{\text{MH}}(\delta) = \left| \frac{N_{\text{acc}}(\delta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\delta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we are maximising the effective sample size as
discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\delta) =
\neff(\delta)$.

@@

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the resulting form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_\Sigma - \frac{1}{2}\big\|u\big\|^2_{\mathcal{T}}\right),
\end{equation}
where $\Sigma$ and $\mathcal{T}$ are the covariances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\Sigma = \sigma^2I_N$ and
$\mathcal{T} = \tau^2I_d$ with $\tau^2 =2$ and $\sigma^2 = 0.1$, and
we observe $u=-2.5$ noisily such that
\[
D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\Sigma).
\]
These values result in a posterior density in which the vast majority
of the density is contained inside the high density region of the
prior. This means that it should be straightforward for the algorithm
to find the stationary distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_Y || \mu_0) = 2.67$ for this problem.

@@

{\bf (1) Finding the optimal parameters}: To find the optimal
parameters we choose 32 values of $\delta$ evenly spaced on a log scale
between $[10^{-5}, 2]$. We run the PAIS-pCNL algorithm for 10,000 iterations and pCNL for 100,000 iterations, each with $M=50$ processes. We took 32 repeats of both algorithms and then used the medians of the statistics to find the optimal parameters.

@@

\subsubsection{Optimal values of $\delta$}\label{sec:Optimal values P1}

@@

\begin{appendix}
\section{The adaptive PAIS algorithm}\label{Sec:App}
